---
title: "Working with data frames in R (1)"
author: "Matt O'Donnell"
date: "Jan 22, 2016"

output:
  html_document:
    theme: united
    toc: yes
    toc_depth: 4
    keep_md: true
---


## What are data frames?

Data frames are tables that hold data that consist of __rows__ and __columns__, just like a spreadsheet. They are most likely the primary way you will access, manipulate and plot your data from while using R.

R includes a series of included data sets that you can find in the datasets library. Type `library(help='datasets')` to see a list of what is there. To get started we'll use the __mtcars__ dataset, which you can find out about by typing `?mtcars`.

> The data was extracted from the 1974 Motor Trend US magazine, and comprises fuel consumption and 10 aspects of automobile design and performance for 32 automobiles (1973–74 models).

Load the data into your workspace with `data(mtcars)`

```{r}
data(mtcars)
head(mtcars)   # this will show you the first 6 rows by default
```

Data frames are tables like a spreadsheet where rows are 'observations' (e.g. subjects, trials, messages) and columns are 'variables' (e.g. measured characteristics, features)

You can also see the whole data by typing `mtcars`. `head` and `tail` will give you the first and last __n__ rows, e.g.

```{r}
head(mtcars,3)  # first three rows
```

```{r}
tail(mtcars,3)  # last three rows
```

### Get the names of the columns in a dataframe

```{r}
colnames(mtcars)
```

### Get the values from a single column

1. by name

```{r}
mtcars$wt
```

2. by using indexing with the syntax <code>DF[row, column]</code> - indexing starts from 1 and leaving either 'row' or 'column' blank will get all values in that row or column

```{r}
mtcars[,1]  # gets the first column
```

which is the same as

```{r}
mtcars$mpg
```

ranges of rows or columns can be specified as:

* continous ranges, e.g. `df[1:5,]` means rows 1,2,3,4&5
* specific indices using c(), e.g. `df[c(1,3,5,11),]` means rows 1,3,5 & 11
* by a combination e.g., `df[c(1:3,7,11,15),]` means rows 1,2,3,7,11&15
* by column or rowname, e.g. `df[,c('mpg','wt)]` means columns 'mpg' and 'wt'


```{r}
mtcars[1:5,5] # get the data from the fifth column for the first five rows
```

* Combining more than one row and more than one column when indexing a data frame will return another data frame. For example, this will get the __mpg__ and __wt__ columns for the 3rd, 5th, 7th and 9th rows of the __mtcars__ data frame.


```{r}
mtcars[c(3,5,7,9),c('mpg','wt')]
```


## Summarizing columns in data frame

#### 1. Specific columns:

```{r}
summary(mtcars$mpg)
```


```{r}
summary(mtcars$gear)
```


#### 2. All columns in data frame:

```{r}
summary(mtcars)
```

## Plotting summary data

#### 1. Histogram

```{r}
hist(mtcars$mpg)
```

#### 2. Pair scatter plot - if your columns are all numeric or categorical you can do a pairwise plot

```{r}
plot(mtcars)
```

## Plotting relationships between variables

#### 1. using plot(x,y) and specifying each column

```{r}
plot(mtcars$wt, mtcars$mpg)
```

#### 2. using with()

```{r}
with(mtcars, plot(wt, mpg))
```

#### 3. using formula syntax - ``plot(y ~ x, data_frame)``

```{r}
plot(mpg ~ wt, mtcars)
```

## Correlations

- ``cor`` will calculate correlation between pairs of vectors

```{r}
cor(mtcars$wt, mtcars$mpg)
```

```{r}
with(mtcars, cor(wt, mpg))
```

- can also do ranges, e.g. ``cor(mtcars[,1:4])`` which does pairwise correlations of the first four columns of the data frame

```{r}
cor(mtcars[,1:4])
```

- or you can use the other methods of indexing discussed above

```{r}
cor(mtcars[,c('wt','mpg','cyl','disp')])
```

- if you just provide the dataframe all numeric columns will be correlated in pair wise fashion

```{r}
cor(mtcars)
```


## Linear regresson

#### Fit a simple regression

Using the same formula syntax ``DV ~ IV`` used above for plots

```{r}
md <- lm(mpg ~ wt, mtcars)

summary(md)
```


#### Fit a multiple regression 

Use ``+`` to add additional IVs

```{r}
md2 <- lm(mpg ~ wt + cyl, mtcars)

summary(md2)
```

#### Examine an interaction

Use ``*`` to interact two IVs

```{r}
md3 <- lm(mpg ~ wt * cyl, mtcars)

summary(md3)
```

#### Compare models 

The ``anova`` function takes two or more fitted models and does an F-test

```{r}

anova(md,md2,md3)

```

#### Add a regression line to a plot

```{r}

plot(mpg ~ wt, mtcars)
abline(md, col='red')

```


## Aggregating data by a categorical variable

Often you will have a grouping or factor variable and want to use that to summarize
other variables by. For example, in the __mtcars__ data frame the cylinder (cyl) variable could be used to group the other variables to calculate the mean.

Here we can get the mean weight for cars with 4, 6 and 8 cylinders. 

```{r}

aggregate(wt ~ cyl, mtcars, mean)    # formula, data_frame, summary function

```

Here we get the standard deviation of miles per gallon grouped by cylinders

```{r}

aggregate(mpg ~ cyl, mtcars, sd)

```

You can also get two or more groupings using the ``+`` syntax in the formula. For example, get the mean of MPG crossed by cylinders and number of gears:

```{r}

aggregate(mpg ~ cyl + gear, mtcars, mean)

```

```{r}

aggregate(cbind(mpg, wt, disp) ~ cyl + gear, mtcars, mean)

```